---
title: "Bayesian GLMM Part9"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(bayesplot)
library(broom.mixed)
library(rstan)
library(patchwork)
library(modelsummary)
library(DHARMa)
source('helperFunctions.R')
```



# Scenario

Once it is established that mass coral bleaching is occurring on the
Great Barrier Reef (GBR), a monitoring team is mobilised in both the
air and water in order to document the scale and extent of the
bleaching. To better understand the causes and consequences of
bleaching, one marine ecologist (lets call him Terry) was interested
in investigating differences in bleaching extent across different reef
habitats. To do so, aerial surveys were partitioned into four habitats
(C - crest, F - flank, L - lower, U - upper).

Bleaching is scored categorically according to the following scale.

| Bleaching Score | 2016 Equates to  |
|-----------------+------------------|
|               0 | No bleaching     |
|               1 | 0-10% bleaching  |
|               2 | 10-30% bleaching |
|               3 | 30-60% bleaching |
|               4 | > 60% bleaching  |

The GBR is very large and the extent of coral bleaching is not uniform
throughout the reef. Hence, Terry wanted to see if the habitat
patterns were similar throughout the GBR or whether they were
dependent on the overall bleaching severity.

# Read in the data

2016 data only

```{r readData, results='markdown', eval=TRUE}
hughes = read_csv('../data/hughes.csv', trim_ws=TRUE)
glimpse(hughes)
```

| REEF   | HABITAT | SECTOR | SCORE |
|--------|---------|--------|-------|
| 09-357 | C       | North  | 4     |
| 09-357 | F       | North  | 4     |
| 09-357 | U       | North  | 3     |
|        |         |        |       |


# Data preparation


# Exploratory Data Analysis

Proportional (cumulative link) odds-ratio models are useful when the latent
(unmeasured) response is recorded on a ordinal (ordered categorical) scale.
When this is the case, we can calculate the probability of a that an observed
ordinal score ($y$) is less than or equal to any given level ($i$: category) given a set of
covariates ($\boldsymbol{X}$) according to the following:

$$ Pr(y\le i|\boldsymbol{X}) =
\frac{1}{1+e^{-(\theta_i - \boldsymbol{\beta}.\boldsymbol{X})}}\\ $$

where $y$ is the observed categorical response, $\boldsymbol{X}$ is a ($n \times
p$) effects model matrix, $\boldsymbol{\beta}$ are the $p$ effects parameters
and $\theta_i$ are the $K-1$ category thresholds




# Fit the model

::: {.panel-tabset}


:::

# MCMC sampling diagnostics

::: {.panel-tabset}


:::

# Model validation 

::: {.panel-tabset}


:::

# Model investigation 

::: {.panel-tabset}


:::




# Further investigations 

::: {.panel-tabset}

:::





