---
title: "Bayesian GLMM Part5"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse) #for data wrangling
library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(knitr)     #for kable
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(brms)
library(broom.mixed)
library(tidybayes)
library(bayesplot)
library(standist)   #for visualizing distributions
library(cmdstanr)   #for cmdstan
library(ggeffects)
library(rstan)
library(DHARMa)
library(ggridges)
library(easystats)     #framework for stats, modelling and visualisation
library(patchwork)
library(geoR) 
source('helperFunctions.R')
```

# Scenario

Some ornithologists were interested in the degree of sibling negotiations in owl
chicks.  Specifically, they wanted to explore how sibling negotiations were
affected by feeding satiety and the sex of the parent returning to the nest.
The ornithologists had accessed to a number of owl nests and were able to count
(via recording equipment) the number of sibling negotiations (calls) that the
owl chicks made when the parent returned to the nest.

We could hypothesise that the chicks might call more if they were hungry.  As
part of the investigation, the researchers were able to provided supplementary
food.  As such, they were able to manipulate the conditions such that sometimes
the chicks in a nest would be considered deprived of supplementary food and at
other times they were satiated.  

As a parent returned, the researchers recorded the number of sibling
negotiations (calls) along with the sex of the parent.  Since the number of
calls is likely to be a function of the number of chicks (the more chicks the
more calls), the researchers also counted the number of siblings in the brood. 

Each nest was measured on multiple occasions.  Hence, we must include the nest
as a random effect to account for the lack of independence between observations
on the same set of siblings.

# Read in the data

```{r readData, results='markdown', eval=TRUE}
owls <- read_csv("../data/owls.csv", trim_ws = TRUE)
```


# Data preparation


# Exploratory data analysis



Random intercept/slope model formula:

$$
\begin{align}
y_i &\sim{} \mathcal{NB}(\mu_{ij}, \phi)\\
ln\left(\mu_{ij}\right) &=\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}\\
\beta_0 &\sim{} \mathcal{N}(0.4, 0.7)\\
\beta_1 &\sim{} \mathcal{N}(0, 1)\\
\boldsymbol{\gamma_j} &\sim{} \mathcal{N}(0, \boldsymbol{\Sigma})\\
\boldsymbol{\Sigma} &= \boldsymbol{D}({\sigma_n^2})\boldsymbol{\Omega}\boldsymbol{D}({\sigma_n^2})\\
\boldsymbol{\Omega} &\sim{} LKJ(\zeta)\\
\sigma_n^2 &\sim{} \mathcal{t}(3,0,0.7)\\
\phi &\sim{} \mathcal{\Gamma}(0.01, 0.01)\\
\end{align}
$$




where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of the fixed
and random effects parameters respectively and $\bf{X}$ is the model matrix
representing the overall intercept and effects of food treatment, sex of parent,
arrival time (and various interactions) on the number of sibling negotiations.
Brood size was also incorporated as an offset.  $\bf{Z}$ represents a cell means
model matrix for the random intercepts associated with individual nests.


# Fit the model 

This model has the potential to be more complex than most of the
models in previous worksheets. Therefore, we will use the current
example to illustrate the process of building a model up gradually
rather than attempting to fit a more complex model from the start. The
benifit of gradually building a model is that it becomes much easier
to construct more sensible and appropriate models that perform nicely
and meet all the various assumptions.

As we intend to build a number of models, we will nest the MCMC
sampling diagnostics and Model validation sections under the Fit the
model section since each model will have its own associated diagnostic
testing.

Typically, when we are building up models, we start by constructing a
model that only contains the hierarchical (random effects) structure.



### Model 2

**Zero-inflated vs hurdle models**

- zero-inflated models: are a mixture of Bernoulli and Poisson (or negative
  binomial) distributions to model situations where it is believed that the
  observed data are the result of two combined processes (a count process that
  governs how many items there are to count and a Bernoulli process that governs
  whether the items are detectable).  In this way, zero-inflated models are
  useful in situations where we believe that some of the zeros are false zeros
  (recorded as zeros when they should not have been).
- hurdle models: are for situations where the response itself is thought to be
  the result of two processes: one that governs whether there is a response and
  then another that governs the value of the response (must be positive).

:::: {.panel-tabset}

#### Fit model


```{r posterior2h_2c, results='markdown', eval=TRUE, fig.width = 7, fig.height = 5}
owls_brm2b |> SUYR_prior_and_posterior()
```


#### MCMC sampling diagnostics 



#### Model validation

::::: {.panel-tabset}

##### pp check




##### DHARMa residuals


:::::

::::

### Model 3

:::: {.panel-tabset}
#### Fit model

```{r posterior2h_3c, results='markdown', eval=TRUE, fig.width = 7, fig.height = 5}
owls_brm3b |> SUYR_prior_and_posterior()
```

#### MCMC sampling diagnostics 



#### Model validation

::::: {.panel-tabset}

##### pp check




##### DHARMa residuals


:::::

::::

### Model 4

:::: {.panel-tabset}
#### Fit model


```{r posterior2h_4c, results='markdown', eval=TRUE, fig.width = 7, fig.height = 5}
owls_brm4b |> SUYR_prior_and_posterior()
```

#### MCMC sampling diagnostics 



#### Model validation

::::: {.panel-tabset}

##### pp check




##### DHARMa residuals


:::::

::::

### Model 5

:::: {.panel-tabset}
#### Fit model

```{r posterior2h_5c, results='markdown', eval=TRUE, fig.width = 7, fig.height = 5}
## owls_brm5b |> SUYR_prior_and_posterior() 
```

#### MCMC sampling diagnostics 



#### Model validation

::::: {.panel-tabset}

##### pp check




##### DHARMa residuals


:::::

::::

### Comparisons


Conclusions:
- evidently, the zero-inflated negative binomial model with random
  intercepts/slopes (owls_brm5b) is the best model...

:::
<!-- END_PRIVATE-->


# Partial effects plots 


# Model investigation 

# Further investigations 


# References
