---
title: "Bayesian GLM Part8"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)
library(brms)
library(dagitty)
library(ggdag)
library(patchwork)
source('helperFunctions.R')
```

# Scenario

To investigate the effects of low light and a herbicide (diuron) on
algae, @King_2022 set up an experiment in which algae were exposed to
different combinations of light levels and diuron concentration. Over
a 72 hour period, chlorophyll-a fluorescence and cell density were
measured.

The data are in the file
**king.csv** in the **data** folder.


| Time | Light | Diuron | Block | t0       | celld     | lncelld     | PI          |
|------|-------|--------|-------|----------|-----------|-------------|-------------|
| 72   | 20    | 0.11   | 2     | 35.24612 | 125.94344 | 4.835832917 | 1.012390192 |
| 72   | 20    | 0.11   | 3     | 39.6347  | 114.65852 | 4.74195832  | 0.929296694 |
| 72   | 20    | 0.11   | 4     | 34.61918 | 167.32148 | 5.119916992 | 0.977363859 |
| 72   | 20    | 0.33   | 2     | 35.4551  | 126.77936 | 4.842448253 | 1.02771364  |
| 72   | 20    | 0.33   | 3     | 45.69512 | 155.20064 | 5.044718731 | 0.95016989  |
| 72   | 20    | 0.33   | 4     | 46.53104 | 162.72392 | 5.092055022 | 0.972060034 |
| 72   | 20    | 1      | 2     | 48.62084 | 165.85862 | 5.111135739 | 0.914524138 |
| \... | \...  | \...   | \...  | \...     | \...      | \...        | \...        |

: Format of the king.csv data file {#tbl-king .table-condensed}

---------------- ---------------------------------------------------
**Time**:        Hour of measurements
**Light**:       Light level treatment (5, 20, 80μmol photons m-2 s-1) - Predictor variable
**Diuron**:      Diuron herbicide concentration (0, 0.11, 0.33, 1, 3μg/L) - Predictor variable
**Block**:       Block - we will ignore this for now
**t0**:          Cell density at time 0
**celld**        Cell density at treatment level - Response variable
**PI**           Photosynthetic inhibition (%), photosynthetic yield (as a % of controls) - Response variable
---------------- ---------------------------------------------------
 
: Description of the variables in the king data file {#tbl-king .table-condensed}

# Read in the data

```{r}
#| label: readData
king <- read_csv("../data/king.csv", trim_ws = TRUE)
```


# Process data

```{r}
#| label: processData
#| eval: false
king <- king |>
    mutate(
      fLight = factor(Light),
      fDiuron = factor(Diuron),
      Block = factor(Block)
    )
```


# Research questions

- Q1: estimate the effect of light on photosynthetic inhibition
- Q2: estimate the effect of photosynthetic inhibition on algal cell density
- Q3: estimate the effect of light on algal cell density
- Q4: estimate the effect of diuron (herbicide) on photosynthetic inhibition
- Q5: estimate the effect of diuron (herbicide) on algal cell density

# Naive approach


# Structural Causal Modelling


1. Create a conceptual model of the system of interest

2. Choose a statistical test

3. Test the consistency of your conceptual model with the data

4. Identify biases (confounding, overcontrol, collider) that you need
   to adjust for when testing the cause of interest

5. Test for causality with the appropriate statistical test

6. Repeat steps 3-5 for each cause of interest

7. Repeat steps 1-6 for other conceptual models

# Step 1 - conceptual model

@King_2022 hypothesised that photosynthetic inhibition would be
effected by light levels and that the diuron herbicide would effect
photosynthesis by blocking certain chemical reactions. In tern,
changes in photosynthesis could have an effect on algal cell density.
Furthermore, diuron might also have a direct effect on cell density by
causing cell damage.

Lets start by creating a DAG that reflects this conceptual model.


# Step 2 - choose a statistical test

For the main inference tests we will fit Bayesian generalized linear
models, however, for the purpose of testing the consistency of our
conceptual model (DAG) with data, we will just use more simple
frequentist analyses.

# Step 3 - test the consistency of the DAG with observed data

Lets now determine what all the implied assumptions (conditional independencies) of our DAG are.


And challenge those assumptions with the data.


# Step 4 - identify biases that need to be adjusted for

Our goal is to be able to investigate the effects of light and diuron
on both algal photosynthesis and cell density. So lets explore each
one separately, starting with photosynthesis.

::: {.panel-tabset}

## Total effect of Light on photosynthesis


## Total Light on cell density

## Total photosynthesis on cell density



## Diuron on cell density


## Diuron on photosynthetic inhibition


:::


**Conclusions**:

- Response: PI
  - Light
    - can include diuron
  - Diuron
    - 
- Response: celld
  - Light
    - total: can include diuron
    - direct: must include PI and diuron
  - PI
    - total: must include diuron
    - total: can also include Light
    - direct: must include diuron
  - Diuron
    - total: can include Light
    - direct: must include PI

# Step 5 - test for causality

::: {.panel-tabset}

## Total effect of Light on photosynthesis

```{r}
#| label: processData
#| results: hide
#| eval: true
#| echo: true
#| cache: false
```

### Exploratory data analysis



## Total effect of Light on cell density

### Exploratory data analysis



## Total effect of photosynthetic inhibition on cell density

### Exploratory data analysis



## Total effect of diuron on cell density

### Exploratory data analysis



:::
