---
title: "Regression trees example 1"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: [default, ../public/resources/ws-style.scss]
    css: ../public/resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, tidy='styler')
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(gbm)         #for gradient boosted models
## library(gbm3)         #for gradient boosted models
library(caret)
library(car)
library(pdp)
library(ggfortify)
library(randomForest)
library(tidyverse)
library(patchwork)
```


# Scenario
@newman-2015 sought to investigate the relative importance of reef habitat
complexity and structural components of the benthos to the total
richness of fish species in Caribbean reefs. To do so, they recorded
they surveyed fish and benthos from one hundred 25m^2 plots within
three marine reserves. Plots were haphazardly selected within three
Countries (St Vincent and the Grenadines, Puerto Rico and Bonaire) and
to represent a range of structural and complexity conditions which
comprised:

- the number of corals
- the number of tall corals
- slope angle
- maximum sponge height
- maximum octocoral height
- reef complexity


Format of newman.csv data file

| total_rich | Reef_complexity | No_tall_corals | No_corals | Sponge_max_height | Octocoral_max_height | Slope_angle | Country |
|------------|-----------------|----------------|-----------|-------------------|----------------------|-------------|---------|
|            |                 |                |           |                   |                      |             |         |
| 25         | 1               | 0              | 14        | 5                 | 13                   | 0           | SVG     |
| 39         | 1               | 0              | 17        | 6                 | 24                   | 0           | SVG     |
| 52         | 1               | 0              | 28        | 12                | 15                   | 0           | SVG     |
| 42         | 1               | 0              | 17        | 11                | 121                  | 5           | SVG     |
| 51         | 1               | 0              | 23        | 13                | 27                   | 5           | SVG     |
| 51         | 1               | 0              | 21        | 38                | 18                   | 0           | SVG     |
| 49         | 1               | 0              | 22        | 31                | 91                   | 5           | SVG     |
| 50         | 1               | 0              | 12        | 14                | 47                   | 5           | SVG     |
| 51         | 1               | 1              | 20        | 34                | 29                   | 0           | SVG     |
| 52         | 2               | 2              | 114       | 0                 | 0                    | 25          | SVG     |

 

# Read in the data

```{r readData, results='markdown', eval=TRUE}
newman <- read_csv('../public/data/newman.csv', trim_ws=TRUE)
glimpse(newman)
```

<!-- START_PRIVATE-->
::: {.panel-tabset}

## glimpse
```{r}
#| label: examinData
#| dependson: readData

newman |> glimpse()
```

## head
```{r}
#| label: headData
#| dependson: readData
## Explore the first 6 rows of the data
newman |> head()
```

## str
```{r}
#| label: strData
#| dependson: readData
newman |> str()
```

## Easystats (datawizard)

```{r}
#| label: easyData
#| dependson: readData
newman |> datawizard::data_codebook()
```
:::

<!-- END_PRIVATE-->

# Exploratory data analysis 

# Fit the model


# Explore relative influence


# Explore partial effects
# Bootstrapping

