---
title: "Bayesian GLM Part8"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: [default, ../public/resources/ws-style.scss]
    css: ../public/resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r}
#| label: setup
#| include: false
#| cache: false
knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries


```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)
library(brms)
library(dagitty)
library(ggdag)
library(patchwork)
source('helperFunctions.R')
```

# Scenario

To investigate the effects of low light and a herbicide (diuron) on
algae, @King_2022 set up an experiment in which algae were exposed to
different combinations of light levels and diuron concentration. Over
a 72 hour period, chlorophyll-a fluorescence and cell density were
measured.

The data are in the file
**king.csv** in the **data** folder.


| Time | Light | Diuron | Block | t0       | celld     | lncelld     | PI          |
|------|-------|--------|-------|----------|-----------|-------------|-------------|
| 72   | 20    | 0.11   | 2     | 35.24612 | 125.94344 | 4.835832917 | 1.012390192 |
| 72   | 20    | 0.11   | 3     | 39.6347  | 114.65852 | 4.74195832  | 0.929296694 |
| 72   | 20    | 0.11   | 4     | 34.61918 | 167.32148 | 5.119916992 | 0.977363859 |
| 72   | 20    | 0.33   | 2     | 35.4551  | 126.77936 | 4.842448253 | 1.02771364  |
| 72   | 20    | 0.33   | 3     | 45.69512 | 155.20064 | 5.044718731 | 0.95016989  |
| 72   | 20    | 0.33   | 4     | 46.53104 | 162.72392 | 5.092055022 | 0.972060034 |
| 72   | 20    | 1      | 2     | 48.62084 | 165.85862 | 5.111135739 | 0.914524138 |
| \... | \...  | \...   | \...  | \...     | \...      | \...        | \...        |

: Format of the king.csv data file {#tbl-king .table-condensed}

---------------- ---------------------------------------------------
**Time**:        Hour of measurements
**Light**:       Light level treatment (5, 20, 80μmol photons m-2 s-1) - Predictor variable
**Diuron**:      Diuron herbicide concentration (0, 0.11, 0.33, 1, 3μg/L) - Predictor variable
**Block**:       Block - we will ignore this for now
**t0**:          Cell density at time 0
**celld**        Cell density at treatment level - Response variable
**PI**           Photosynthetic inhibition (%), photosynthetic yield (as a % of controls) - Response variable
---------------- ---------------------------------------------------
 
: Description of the variables in the king data file {#tbl-king .table-condensed}

# Read in the data

```{r}
#| label: readData
king <- read_csv("../public/data/king.csv", trim_ws = TRUE)
```

<!-- START_PRIVATE-->
::: {.panel-tabset}

## glimpse

```{r}
#| label: examinData
glimpse(king)
```

## head

```{r}
## Explore the first 6 rows of the data
head(king)
```

## str

```{r}
str(king)
```

## Easystats (datawizard)

```{r}
king |> datawizard::data_codebook()
```

:::

<!-- END_PRIVATE-->

# Process data

```{r}
#| label: processData
#| eval: false
king <- king |>
    mutate(
      fLight = factor(Light),
      fDiuron = factor(Diuron),
      Block = factor(Block)
    )
```


# Research questions

- Q1: estimate the effect of light on photosynthetic inhibition
- Q2: estimate the effect of photosynthetic inhibition on algal cell density
- Q3: estimate the effect of light on algal cell density
- Q4: estimate the effect of diuron (herbicide) on photosynthetic inhibition
- Q5: estimate the effect of diuron (herbicide) on algal cell density


# Naive approach

<!-- START_PRIVATE-->

```{r}
#| label: eda_mod0a
#| results: hide
#| eval: true
#| echo: true
#| cache: true
#| fig-width: 13
#| fig-height: 4
g1 <-
    king |>
    ggplot(aes(y = PI, x = Diuron, colour = factor(Light))) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_classic()
g2 <- 
  king |>
  ggplot(aes(y = celld, x = PI, colour = factor(Light))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()
g3 <- 
  king |>
  ggplot(aes(y = celld, x = Diuron, colour = factor(Light))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()
g1 + g2 + g3 + plot_layout(guides = "collect")
```

::: {.panel-tabset}

## Response: PI



```{r}
#| label: brms_mod0a
#| eval: true
#| echo: true
#| cache: true
form <- bf(PI ~ scale(Diuron) * factor(Light))
## summarise data to help inform priors
king |>
    group_by(Light) |>
    summarise(PI_mu = median(PI), PI_sd = sd(PI), PI_mad = mad(PI))

## define priors
priors <- prior(normal(1.4, 0.3), class = "Intercept") +
  prior(normal(0, 0.5), class = "b") +
  prior(student_t(3, 0, 0.3), class = "sigma")
## fit the mode
king_brms0a <- brm(
  form,
  prior = priors,
  data = king,
  refresh = 0,
  backend = "cmdstanr"
)
```

```{r}
#| label: brms_mod0a_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms0a)
```

**Conclusions:**

- increasing diuron concentration associated with decreasing photosynthetic inhibition ()
- increasing light is associated with decreasing photosynthetic inhibition 


## Response: cell density

```{r}
#| label: brms_mod0b
#| eval: true
#| echo: true
#| cache: true
form <- bf(celld ~ scale(Diuron) + scale(PI) + Light)
## summarise data to help inform priors
king |>
    group_by(Light) |>
    summarise(celld_mu = median(celld), celld_sd = sd(celld), celld_mad = mad(celld))

## define priors
priors <- prior(normal(150, 50), class = "Intercept") +
  prior(normal(0, 20), class = "b") +
  prior(student_t(3, 0, 50), class = "sigma")
## fit the mode
king_brms0b <- brm(
  form,
  prior = priors,
  data = king |> mutate(Light = factor(Light)),
  refresh = 0,
  backend = "cmdstanr"
)
```
 
```{r}
#| label: brms_mod0b_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms0b)
```

**Conclusions:**

- no effect of diuron concentration on algal cell density
- no effect of photosynthetic inhibition on algal cell density
- no effect of light on algal cell density

:::
<!-- END_PRIVATE-->

# Structural Causal Modelling


1. Create a conceptual model of the system of interest

2. Choose a statistical test

3. Test the consistency of your conceptual model with the data

4. Identify biases (confounding, overcontrol, collider) that you need
   to adjust for when testing the cause of interest

5. Test for causality with the appropriate statistical test

6. Repeat steps 3-5 for each cause of interest

7. Repeat steps 1-6 for other conceptual models

# Step 1 - conceptual model

@King_2022 hypothesised that photosynthetic inhibition would be
effected by light levels and that the diuron herbicide would effect
photosynthesis by blocking certain chemical reactions. In tern,
changes in photosynthesis could have an effect on algal cell density.
Furthermore, diuron might also have a direct effect on cell density by
causing cell damage.

Lets start by creating a DAG that reflects this conceptual model.

<!-- START_PRIVATE-->
```{r}
#| label: dag
#| eval: true
#| cache: false
king_dag <- dagify(
    celld ~ PI + Diuron,
    PI ~ Light + Diuron,
    exposure =  "Light",
    outcome = "celld"
)
ggdag(king_dag, text = TRUE, text_size = 2.5) +
  theme_dag_blank()
```
<!-- END_PRIVATE-->

# Step 2 - choose a statistical test

For the main inference tests we will fit Bayesian generalized linear
models, however, for the purpose of testing the consistency of our
conceptual model (DAG) with data, we will just use more simple
frequentist analyses.

# Step 3 - test the consistency of the DAG with observed data

Lets now determine what all the implied assumptions (conditional independencies) of our DAG are.

<!-- START_PRIVATE-->
```{r}
#| label: implied conditional independencies
#| eval: true
#| cache: false
dagitty::impliedConditionalIndependencies(king_dag) 
```

**Conclusions**:

- Diuron is independent of Light
- Light is independent of celld when conditioned on both Diuron and PI

<!-- END_PRIVATE-->

And challenge those assumptions with the data.

<!-- START_PRIVATE-->
```{r}
#| label: local tests
#| eval: true
#| cache: false
tests <- localTests(x = king_dag, data = king)
tests
```

**Conclusions**:

- it seems that our conceptual model is consistent with the data.

<!-- END_PRIVATE-->

# Step 4 - identify biases that need to be adjusted for

Our goal is to be able to investigate the effects of light and diuron
on both algal photosynthesis and cell density. So lets explore each
one separately, starting with photosynthesis.

::: {.panel-tabset}

## Total effect of Light on photosynthesis

<!-- START_PRIVATE-->
```{r}
#| label: ggdag paths 1
#| eval: true
#| cache: false
#| fig-width: 7
#| fig-height: 5
ggdag_paths(king_dag, from = "Light", to = "PI", text_col = "black", shadow = TRUE) +
    theme_dag_blank(panel.border = element_rect(fill = NA)) 
```

**Conclusions**:

- there is a single direct path

```{r}
#| label: adjustment sets 1a
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "PI",
  type = "minimal",
  effect = "total")
```

We can also ask whether there are any other optional covariates that
we could include without introducing biases.

```{r}
#| label: adjustment sets 1b
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "PI",
  type = "canonical",
  effect = "total")
```

- we could also include diuron

```{r}
#| label: adjustment sets 1c
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "PI",
  type = "minimal",
  effect = "direct")
```

<!-- END_PRIVATE-->

## Total Light on cell density

<!-- START_PRIVATE-->
```{r}
#| label: ggdag paths 2
#| eval: true
#| cache: false
#| fig-width: 7
#| fig-height: 5
ggdag_paths(king_dag, from = "Light", to = "celld", text_col = "black", shadow = TRUE) +
    theme_dag_blank(panel.border = element_rect(fill = NA)) 
```

**Conclusions**:

- there is a single path
  - a pipe (`Light -> PI -> celld`)

```{r}
#| label: adjustment sets 2a
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "celld",
  type = "minimal",
  effect = "total")
```

- to explore the total effects of Light on celld, we do not condition on anything

We can also ask whether there are any other optional covariates that
we could include without introducing biases.

```{r}
#| label: adjustment sets 2b
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "celld",
  type = "canonical",
  effect = "total")
```

- we could also include diuron

The conceptual model does not include a direct pathway from Light to cell density,
but if we wanted to examine this, what would we need to condition on?

```{r}
#| label: adjustment sets 2c
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Light", 
  outcome = "celld",
  type = "minimal",
  effect = "direct")
```

- we would need to include both Diuron and PI

<!-- END_PRIVATE-->
## Total photosynthesis on cell density

<!-- START_PRIVATE-->
```{r}
#| label: ggdag paths 4
#| eval: true
#| cache: false
#| fig-width: 7
#| fig-height: 5
ggdag_paths(king_dag, from = "PI", to = "celld", text_col = "black", shadow = TRUE) +
    theme_dag_blank(panel.border = element_rect(fill = NA)) 
```

**Conclusions**:

- there is are two paths
  - direct (`PI -> celld`)
  - a forl (`PI <- Diuron -> celld`) 
    - this is a backdoor
    - we want to block this path

```{r}
#| label: adjustment sets 4a
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "PI", 
  outcome = "celld",
  type = "minimal",
  effect = "total")
```

- to explore the total effects of photosynthesis inhibition on cell
  density, we **must** condition on diuron

We can also ask whether there are any other optional covariates that
we could include without introducing biases.

```{r}
#| label: adjustment sets 4b
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "PI", 
  outcome = "celld",
  type = "canonical",
  effect = "total")
```

- we could also include light, but we already know that we must
  include diuron

The conceptual model also includes a direct pathway from PI to cell density,
so if we wanted to examine this, what would we need to condition on?

```{r}
#| label: adjustment sets 4c
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "PI", 
  outcome = "celld",
  type = "minimal",
  effect = "direct")
```

- again, as a minimum, we **must** condition on diuron

<!-- END_PRIVATE-->


## Diuron on cell density

<!-- START_PRIVATE-->
```{r}
#| label: ggdag paths 3
#| eval: true
#| cache: false
#| fig-width: 12
#| fig-height: 5
ggdag_paths(king_dag, from = "Diuron", to = "celld", text_col = "black", shadow = TRUE) +
    theme_dag_blank(panel.border = element_rect(fill = NA)) 
```

**Conclusions**:

- there are two paths
  - one is a pipe (`Diron -> PI -> celld`)
  - one is direct (`Diuron -> celld`)

```{r}
#| label: adjustment sets 3a
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "celld",
  type = "minimal",
  effect = "total")
```

- to explore the total effects of Diuron on cell density, we do not
  condition on any other covariate


We can also ask whether there are any other optional covariates that
we could include without introducing biases.

```{r}
#| label: adjustment sets 3b
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "celld",
  type = "canonical",
  effect = "total")
```

- we could also include Light

If we wanted to examine the direct effects of Diuron on cell density,
what would we need to condition on?

```{r}
#| label: adjustment sets 3c
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "celld",
  type = "minimal",
  effect = "direct")
```

- we would need to include PI

<!-- END_PRIVATE-->

## Diuron on photosynthetic inhibition

<!-- START_PRIVATE-->
```{r}
#| label: ggdag paths 5
#| eval: true
#| cache: false
#| fig-width: 12
#| fig-height: 5
ggdag_paths(king_dag, from = "Diuron", to = "PI", text_col = "black", shadow = TRUE) +
    theme_dag_blank(panel.border = element_rect(fill = NA)) 
```

**Conclusions**:

- there are two paths
  - one is a pipe (`Diron -> PI -> celld`)
  - one is direct (`Diuron -> celld`)

```{r}
#| label: adjustment sets 5a
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "PI",
  type = "minimal",
  effect = "total")
```

- to explore the total effects of Diuron on cell density, we do not
  condition on any other covariate


We can also ask whether there are any other optional covariates that
we could include without introducing biases.

```{r}
#| label: adjustment sets 5b
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "PI",
  type = "canonical",
  effect = "total")
```

- we could also include Light

If we wanted to examine the direct effects of Diuron on cell density,
what would we need to condition on?

```{r}
#| label: adjustment sets 5c
#| eval: true
#| cache: false
adjustmentSets(king_dag,
  exposure = "Diuron", 
  outcome = "PI",
  type = "minimal",
  effect = "direct")
```

- we would need to include PI

<!-- END_PRIVATE-->

:::


<!-- START_PRIVATE-->
**Conclusions**:

- Response: PI
  - Light
    - can include diuron
  - Diuron
    - can also have light 
- Response: celld
  - Light
    - total: can include diuron
    - direct: must include PI and diuron
  - PI
    - total: must include diuron
    - total: can also include Light
    - direct: must include diuron
  - Diuron
    - total: can include Light
    - direct: must include PI

PI: ~ Light + Diuron
Celld (total): ~ Light + Diuron

<!-- END_PRIVATE-->

# Step 5 - test for causality

::: {.panel-tabset}

## Total effect of Light on photosynthesis

<!-- START_PRIVATE
```{r}
#| label: processData
#| results: hide
#| eval: true
#| echo: true
#| cache: false
```

### Exploratory data analysis


```{r}
#| label: eda_mod1a
#| results: hide
#| eval: true
#| echo: true
#| cache: true
king |>
    ggplot(aes(y = PI, x = Diuron, fill = fLight)) +
    geom_boxplot() +
    theme_classic()
```
 
```{r}
#| label: brms_mod1a
#| eval: true
#| echo: true
#| cache: true
form <- bf(PI ~ fLight * scale(Diuron))
## summarise data to help inform priors
king |>
    group_by(fLight) |>
    summarise(PI_mu = median(PI), PI_sd = sd(PI), PI_mad = mad(PI))

## define priors
priors <- prior(normal(1.4, 0.3), class = "Intercept") +
  prior(normal(0, 0.5), class = "b") +
  prior(student_t(3, 0, 0.3), class = "sigma")
## fit the mode
king_brms1a <- brm(
  form,
  prior = priors,
  data = king,
  refresh = 0,
  backend = "cmdstanr"
)
```

```{r}
#| label: brms_mod1a_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms1a)
```

**Conclusions:** 

- increasing light reduces photosynthetic inhibition
- increasing diuron reduces photosynthetic inhibition

<!-- END_PRIVATE-->

## Total effect of Light on cell density

<!-- START_PRIVATE-->
### Exploratory data analysis


```{r}
#| label: eda_mod2a
#| results: hide
#| eval: true
#| echo: true
#| cache: true
king |>
    ggplot(aes(y = celld, x = Diuron, fill = fLight)) +
    geom_boxplot() +
    theme_classic()
```
 
```{r}
#| label: brms_mod2a
#| eval: true
#| echo: true
#| cache: true
form <- bf(celld ~ fLight * scale(Diuron))
## summarise data to help inform priors
king |>
    group_by(fLight) |>
    summarise(celld_mu = median(celld), celld_sd = sd(celld), celld_mad = mad(celld))

## define priors
priors <- prior(normal(150, 50), class = "Intercept") +
  prior(normal(0, 30), class = "b") +
  prior(student_t(3, 0, 50), class = "sigma")
## fit the mode
king_brms2a <- brm(
  form,
  prior = priors,
  data = king,
  refresh = 0,
  backend = "cmdstanr"
)
```

```{r}
#| label: brms_mod2a_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms2a)
```

**Conclusions:** 

- increasing light increases algal cell density (although only at 20 units)
- increasing diuron concentration reduces algal cell density
- however, there is an interaction between diuron and light levels

<!-- END_PRIVATE-->

## Total effect of photosynthetic inhibition on cell density

<!-- START_PRIVATE-->
### Exploratory data analysis


```{r}
#| label: eda_mod3a
#| results: hide
#| eval: true
#| echo: true
#| cache: true
king |>
    ggplot(aes(y = celld, x = PI, fill = fLight)) +
    geom_boxplot() +
    theme_classic()
```
 
```{r}
#| label: brms_mod3a
#| eval: true
#| echo: true
#| cache: true
form <- bf(celld ~ scale(PI) * fLight * scale(Diuron))
## summarise data to help inform priors
king |>
    group_by(fLight) |>
    summarise(celld_mu = median(celld), celld_sd = sd(celld), celld_mad = mad(celld))

## define priors
priors <- prior(normal(150, 50), class = "Intercept") +
  prior(normal(0, 30), class = "b") +
  prior(student_t(3, 0, 50), class = "sigma")
## fit the mode
king_brms3a <- brm(
  form,
  prior = priors,
  data = king,
  refresh = 0,
  backend = "cmdstanr"
)
```

```{r}
#| label: brms_mod3a_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms3a)
```

**Conclusions:** 

- no evidence that photosynthetic inhibition has an effect on algal cell density
- but ignore diuron and light

<!-- END_PRIVATE-->

## Total effect of diuron on cell density

<!-- START_PRIVATE-->
### Exploratory data analysis

```{r}
#| label: eda_mod4a
#| results: hide
#| eval: true
#| echo: true
#| cache: true
king |>
    ggplot(aes(y = celld, x = Diuron, fill = fLight)) +
    geom_boxplot() +
    theme_classic()
```
 
```{r}
#| label: brms_mod4a
#| eval: true
#| echo: true
#| cache: true
form <- bf(celld ~ fLight * scale(Diuron))
## summarise data to help inform priors
king |>
    group_by(fLight) |>
    summarise(celld_mu = median(celld), celld_sd = sd(celld), celld_mad = mad(celld))

## define priors
priors <- prior(normal(150, 50), class = "Intercept") +
  prior(normal(0, 30), class = "b") +
  prior(student_t(3, 0, 50), class = "sigma")
## fit the mode
king_brms4a <- brm(
  form,
  prior = priors,
  data = king,
  refresh = 0,
  backend = "cmdstanr"
)
```

```{r}
#| label: brms_mod4a_sum
#| results: markup
#| eval: true
#| echo: true
#| cache: false
summary(king_brms4a)
```

**Conclusions:** 

- increasing light levels increase algal cell density
- increasing diuron concentration decreases algal cell density
- there is evidence of an interaction between light and diuron concentration

<!-- END_PRIVATE-->


:::
