---
title: "Bayesian GLMM example 7"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: spacelab
    css: ../public/resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    highlight-style: zenburn
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(broom.mixed)
library(rstan)
library(patchwork)
library(DHARMa)
```

<!-- START_PRIVATE-->
# Original data

```{r readDataP, results='markdown', eval=TRUE}
hughes <- read_csv('../public/data/hughes_full.csv', trim_ws=TRUE)
glimpse(hughes)
```

| Bleaching Score | 2016 Equates to  | 2017 Equates to  |
|-----------------+------------------+------------------|
|               0 | No bleaching     | No bleaching     |
|               1 | 0-10% bleaching  | 0-10% bleaching  |
|               2 | 10-30% bleaching | 10-30% bleaching |
|               3 | 30-60% bleaching | 30-60% bleaching |
|               4 | > 60% bleaching  | 60-80% bleaching |
|               5 | -                | > 80% bleaching  |

# Data processing

- Make a categorical version of Year
- Combine Scores 4 and 5 into a single Score of 4 to be consistent across years
- Make Score an ordered categorical variable
- Generate a numeric version of categorical Score (as required for the `ocat` family.  Note this must be a 1 indexed numeric.
- Relevel Sector such that it is in the order North, Central, South
- Ensure that there is a categorical version of Reef
- Generate a numeric version of the categorical Reef 


```{r processDataP, results='markdown', eval=TRUE}
hughes <- hughes %>%
    mutate(fYear=factor(Year),
           Score=ifelse(Score==5,4,Score),
           oScore = factor(Score, ordered=TRUE),
           nScore = as.numeric(factor(Score, ordered=TRUE)),
           SectorThree=factor(SectorThree, levels=c('North','Central','South')),
           fReef=factor(ReefID),
           nReef=as.numeric(fReef))
# now make a version that is just 2016
hughes <- hughes %>% filter(fYear==2016) %>%
    dplyr::select(REEF=ReefID, HABITAT=Habitat, SECTOR=SectorThree, SCORE=Score)
write_csv(hughes, file='../public/data/hughes.csv')
## hughes.colors = c('#FFFFFF', rev(heat.colors(length(levels(hughes$oScore))))[-1])
```

<!-- END_PRIVATE-->
