---
title: "Bayesian GLMM example 7"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format: 
  html:
    ## Format
    theme: spacelab
    css: ../public/resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    highlight-style: zenburn
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../public/resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache.lazy = FALSE,
                      tidy = "styler")
options(tinytex.engine = "xelatex")
```


# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(car)       #for regression diagnostics
library(broom)     #for tidy output
library(ggfortify) #for model diagnostics
library(sjPlot)    #for outputs
library(knitr)     #for kable
library(effects)   #for partial effects plots
library(emmeans)   #for estimating marginal means
library(MASS)      #for glm.nb
library(MuMIn)     #for AICc
library(tidyverse) #for data wrangling
library(brms)
library(tidybayes)
library(broom.mixed)
library(rstan)
library(patchwork)
library(DHARMa)
```

# Scenario

@Mitchell-2019 presented a data set in which male guppy fish where
placed individually into long, narrow and shallow tanks that
predominately only permitted the fish to swim in two dimensions. The
activity rate (cumulative distance swam) of each guppy fish was
remotely recorded via tracking software during 20 minute trials
conducted each day for 14 days. The order in which individual guppy
fish were recorded each day was randomised and the time of day of each
trial was recorded.

The researchers were interested in whether the the guppy fish become
habituated to the experimental conditions and altered their activity
patterns over time.

The data are in the file
**mitchell.csv** in the **data** folder.

| ID   | TOD  | Day | Dist_moved |
|------|------|-----|------------|
| A39  | 0.39 | 7   | 3439.93    |
| A3   | 0.39 | 7   | 2795.08    |
| A13  | 0.39 | 7   | 5258.79    |
| \... | \... |     |            |

: Format of the mitchell.csv data file {#tbl-mitchell .table-condensed}

---------------- ---------------------------------------------------
**ID**:           Individual guppy fish ID - Random variable
**TOD**:          Time of day of observation as a proportion - Predictor variable
**Day**:          Day post entry into tank - Predictor variable
**Dist_moved**:   Total distance (cm) moved in 20 min trial - Response variable
---------------- ---------------------------------------------------
 
: Description of the variables in the mitchell data file {#tbl-mitchell1 .table-condensed}
 
# Read in the data

```{r readData, results='markdown', eval=TRUE}
mitchell <- read_csv("../public/data/mitchell.csv", trim_ws = TRUE)
```

<!-- START_PRIVATE-->
::: {.panel-tabset}

## glimpse
```{r}
#| label: examinData
glimpse(mitchell)
```

## head
```{r}
## Explore the first 6 rows of the data
head(mitchell)
```

## str
```{r}
str(mitchell)
```

## Easystats (datawizard)
```{r}
mitchell |> datawizard::data_codebook()
```
:::
<!-- END_PRIVATE-->

# Data preparation
<!-- START_PRIVATE--> Let start by declaring the categorical variables
and random effect as factors.


```{r dataProcessing, results='markdown', eval=TRUE, mhidden=TRUE}
mitchell <- mitchell |>
  mutate(ID = factor(ID))
```
<!-- END_PRIVATE-->

# Exploratory data analysis

<!-- START_PRIVATE--> As the response represents a strictly positive
real number (total distance moved), it would make sense to start by
considering a Gamma model.

Note, also that the response variable are relatively large distances
(as the units of distance used were centimeters). It might be worth
dividing these values by 100 or even 1000 during the modelling
process.

<!-- END_PRIVATE-->
Model formula:
$$
y_i/1000 \sim{} \mathcal{Gamma}(\lambda_i)\\
ln(\lambda_i) =\boldsymbol{\beta} \bf{X_i} + \boldsymbol{\gamma} \bf{Z_i}
$$

where $\boldsymbol{\beta}$ and $\boldsymbol{\gamma}$ are vectors of
the fixed and random effects parameters respectively and $\bf{X}$ is
the model matrix representing the overall intercept and effects of day
and time of day on the total distance moved during the trial. $\bf{Z}$
represents a cell means model matrix for the random intercepts
associated with individual guppies.

<!-- START_PRIVATE-->
::: {.panel-tabset}
## Boxplots
```{r eda0, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=7, fig.height=5}
mitchell |> ggplot(aes(y = Dist_moved)) +
  geom_boxplot()
mitchell |> ggplot(aes(y = Dist_moved)) +
  geom_boxplot() +
  scale_y_log10()
mitchell |> ggplot(aes(y = TOD)) +
  geom_boxplot() +
  scale_y_log10()
mitchell |> ggplot(aes(y = Day)) +
  geom_boxplot() 
```
## Scatter plots
```{r eda1, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=7, fig.height=5}
mitchell |> ggplot(aes(y = Dist_moved, x = Day)) +
  geom_point() +
  geom_line(aes(group = ID))
mitchell |> ggplot(aes(y = Dist_moved, x = TOD)) +
  geom_point() +
  geom_line(aes(group = ID))
```
## Facet plots
```{r eda2, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=15, fig.height=15}
mitchell |> ggplot(aes(y = Dist_moved, x = Day)) +
  geom_point() +
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method =  "lm") +
  facet_wrap(~ID)
mitchell |> ggplot(aes(y = Dist_moved, x = TOD)) +
  geom_point() +
  scale_y_log10() +
  scale_x_log10() +
  geom_smooth(method =  "lm") +
  facet_wrap(~ID)
```
:::
<!-- END_PRIVATE-->


# Fit the model 

<!-- START_PRIVATE-->
::: {.panel-tabset}
## brms 
:::: {.panel-tabset}
### Using default priors

In `brms`, the default priors are designed to be weakly informative.  They are
chosen to provide moderate regularisation (to help prevent over fitting) and
help stabilise the computations.

Unlike `rstanarm`, `brms` models must be compiled before they start sampling.
For most models, the compilation of the stan code takes around 45 seconds.

```{r fitModel2a, results='markdown', eval=TRUE, mhidden=TRUE, paged.print=FALSE, tidy.opts = list(width.cutoff = 80)}
mitchell.form <- bf(I(Dist_moved/1000) ~ Day + TOD + (1|ID),
                family=Gamma(link='log'))
options(width=150)
mitchell.form |> get_prior(data = mitchell)
options(width=80)
```

### Defining priors

The following link provides some guidance about defining priors.
[https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations]

When defining our own priors, we typically do not want them to be scaled.

If we wanted to define our own priors that were less vague, yet still not likely
to bias the outcomes, we could try the following priors (mainly plucked out of
thin air):

- $\beta_0$: normal centred at 134 with a standard deviation of 65
  - mean of 134: since `median(log(mitchell$Dist_moved/1000))`
  - sd pf 0.7: since `mad(log(mitchell$Dist_moved/1000))`
- $\beta_1$: normal centred at 0 with a standard deviation of 2
  - sd of 1: since `mad(log(Dist_moved/1000))/mad(scale(Day))`
- $\sigma$: half-t centred at 0 with a standard deviation of 65 OR
  - sd pf 65: since `mad(log(mitchell$Dist_moved/1000))`
- $\sigma$: gamma with shape parameters of 2 and 1
- $\beta_0$: normal centred at 1.5 with a standard deviation of 1.5
  - mean of 1.5: since `mean(log(mitchell$Dist_moved/1000))` or `mean(asinh(mitchell$Dist_moved/(2*1))/log(exp(1)))`
  - sd of 1.5: since `sd(log(mitchell$Dist_moved/1000))` or `sd(asinh(mitchell$Dist_moved/(2*1))/log(exp(1)))`

```{r fitModel2h, results='markdown', eval=TRUE, mhidden=TRUE, cache=FALSE}
mitchell |> 
  summarise(
    Int_mu = median(log(Dist_moved/1000)),
    Int_sd = mad(log(Dist_moved/1000)),
    b1_sd = mad(log(Dist_moved/1000))/mad(scale(Day)),
    b2_sd = mad(log(Dist_moved/1000))/mad(scale(TOD))
    )
standist::visualize("normal(1.8,5)", xlim=c(0,20))
standist::visualize("student_t(3, 0, 2.5)",
                    "cauchy(0,1)",
                    xlim=c(-10,25))
```

```{r fitModel2h1, results='markdown', eval=FALSE, mhidden=TRUE}
priors <- prior(normal(0.4, 0.7), class = 'Intercept') +
    prior(normal(0, 2), class = 'b', coef = 'FoodTreatmentSatiated') +
    prior(normal(0, 2.2), class = 'b', coef = 'SexParentMale') +
    prior(normal(0, 2), class = 'b') +
    prior(student_t(3, 0, 2), class = 'sd') 
    prior(cauchy(0,2), class = 'sd') 

priors <- prior(normal(1.8, 2), class = 'Intercept') +
    prior(normal(0, 2.2), class = 'b', coef = 'FoodTreatmentSatiated') +
    prior(normal(0, 2.2), class = 'b', coef = 'SexParentMale') +
    prior(normal(0, 1), class = 'b') +
    prior(cauchy(0,2), class = 'sd') 
owls.form <- bf(NCalls ~ FoodTreatment*SexParent +
                    offset(log(BroodSize)) + (1|Nest),
                family=poisson(link='log'))
owls.brm2 <- brm(owls.form, 
                 data = owls,
                 prior = priors,
                 sample_prior = 'only',
                 iter = 5000,
                 warmup =2500,
                 chains = 3,  
                 cores = 3,
                 thin = 10,
                 refresh = 0,
                 seed = 123,
                 control =  list(adapt_delta = 0.99),
                 backend = "cmdstanr"
                 )

```

```{r partialPlot2h1a, results='markdown', eval=FALSE, mhidden=TRUE, fig.width=8, fig.height=5}
owls.brm2 |>
    ggpredict(~FoodTreatment*SexParent) |>
    plot(add.data = TRUE)
owls.brm2 |>
    conditional_effects("FoodTreatment:SexParent") |>
    plot(add.data = TRUE)
owls.brm2 |>
    conditional_effects("FoodTreatment:SexParent") |>
  plot(points = TRUE) |>
  _[[1]] +
  scale_y_continuous(trans = scales::pseudo_log_trans())
```

The above seem sufficiently wide whilst at the same time not providing any encouragement for the sampler
to wander off into very unsupported areas.

```{r fitModel2h1b, results='markdown', eval=FALSE, mhidden=TRUE}
owls.brm3 <- update(owls.brm2,  
                       sample_prior = 'yes',
                       iter =  10000,
                       warmup =  5000,
                       control = list(adapt_delta = 0.99),
                       backend =  "cmdstanr",
                       refresh = 0)
save(owls.brm3, file = '../ws/testing/owls.brm3')
```

```{r partialPlot2h1b, results='markdown', eval=FALSE, mhidden=TRUE, fig.width=8, fig.height=5}
owls.brm3 |>  
    ggpredict(~FoodTreatment*SexParent) |>
    plot(add.data = TRUE)
owls.brm3 |>  
    conditional_effects("FoodTreatment:SexParent") |>
    plot(points = TRUE) |> 
  _[[1]] +
  scale_y_continuous(trans = scales::pseudo_log_trans())
```

::::

:::
<!-- END_PRIVATE-->
