---
title: "Regression trees example 1"
author: "Murray Logan"
date: today
date-format: "DD/MM/YYYY"
format:
  html:
    ## Format
    theme: [default, ../resources/ws-style.scss]
    css: ../resources/ws_style.css
    html-math-method: mathjax
    ## Table of contents
    toc: true
    toc-float: true
    ## Numbering
    number-sections: true
    number-depth: 3
    ## Layout
    page-layout: full
    fig-caption-location: "bottom"
    fig-align: "center"
    fig-width: 4
    fig-height: 4
    fig-dpi: 72
    tbl-cap-location: top
    ## Code
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    ## Execution
    execute:
      echo: true
      cache: true
    ## Rendering
    embed-resources: true
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
output_dir: "docs"
documentclass: article
fontsize: 12pt
mainfont: Arial
mathfont: LiberationMono
monofont: DejaVu Sans Mono
classoption: a4paper
bibliography: ../resources/references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE, tidy='styler')
```

# Preparations

Load the necessary libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false

library(gbm)         #for gradient boosted models
## library(gbm3)         #for gradient boosted models
library(car)
library(pdp)
library(ggfortify)
library(randomForest)
library(tidyverse)
library(patchwork)
```


# Scenario
@newman-2015 sought to investigate the relative importance of reef habitat
complexity and structural components of the benthos to the total
richness of fish species in Caribbean reefs. To do so, they recorded
they surveyed fish and benthos from one hundred 25m^2 plots within
three marine reserves. Plots were haphazardly selected within three
Countries (St Vincent and the Grenadines, Puerto Rico and Bonaire) and
to represent a range of structural and complexity conditions which
comprised:

- the number of corals
- the number of tall corals
- slope angle
- maximum sponge height
- maximum octocoral height
- reef complexity


Format of newman.csv data file

| total_rich | Reef_complexity | No_tall_corals | No_corals | Sponge_max_height | Octocoral_max_height | Slope_angle | Country |
|------------|-----------------|----------------|-----------|-------------------|----------------------|-------------|---------|
|            |                 |                |           |                   |                      |             |         |
| 25         | 1               | 0              | 14        | 5                 | 13                   | 0           | SVG     |
| 39         | 1               | 0              | 17        | 6                 | 24                   | 0           | SVG     |
| 52         | 1               | 0              | 28        | 12                | 15                   | 0           | SVG     |
| 42         | 1               | 0              | 17        | 11                | 121                  | 5           | SVG     |
| 51         | 1               | 0              | 23        | 13                | 27                   | 5           | SVG     |
| 51         | 1               | 0              | 21        | 38                | 18                   | 0           | SVG     |
| 49         | 1               | 0              | 22        | 31                | 91                   | 5           | SVG     |
| 50         | 1               | 0              | 12        | 14                | 47                   | 5           | SVG     |
| 51         | 1               | 1              | 20        | 34                | 29                   | 0           | SVG     |
| 52         | 2               | 2              | 114       | 0                 | 0                    | 25          | SVG     |



# Read in the data

```{r readData, results='markdown', eval=TRUE}
newman <- read_csv('../data/newman.csv', trim_ws=TRUE)
glimpse(newman)
```

<!-- START_PRIVATE-->
::: {.panel-tabset}

## glimpse
```{r}
#| label: examinData
#| dependson: readData

newman |> glimpse()
```

## head
```{r}
#| label: headData
#| dependson: readData
## Explore the first 6 rows of the data
newman |> head()
```

## str
```{r}
#| label: strData
#| dependson: readData
newman |> str()
```

## Easystats (datawizard)

```{r}
#| label: easyData
#| dependson: readData
newman |> datawizard::data_codebook()
```
:::

<!-- END_PRIVATE-->

# Exploratory data analysis

<!-- START_PRIVATE-->

```{r}
#| label: process data
#| results: markup
#| eval: true
#| echo: true
#| cache: false
names(newman) <- str_replace_all(names(newman), " ", "_")
newman <- newman |>
  mutate(
    fCountry = factor(Country),
    nCountry = as.numeric(fCountry),
    fReef_complexity = factor(Reef_complexity)
  )
```



::: {.panel-tabset}
## Scatterplot matrix

A scatterplot matrix plots each variable against each other.  It is useful to
provide boxplots in the diagonals to help explore the distributions.


```{r EDA, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=8, fig.height=8}
#| dependson: readData
scatterplotMatrix(~total_rich + Reef_complexity + No_tall_corals +
                    No_corals + Sponge_max_height + Octocoral_max_height +
                    Slope_angle + nCountry, data = newman,
                    diagonal = list(method='boxplot'))
```

**Conclusions:**

- on the top row, total_rich is on the y-axis of each plot
- the second plot from the top left has Reef_complexity on the x-axis
- we mainly want to pay attention to the polarity of any trends to use
  for monotonic specifications

:::


<!-- END_PRIVATE-->

# Fit the model

<!-- START_PRIVATE-->
::: {.panel-tabset}

Now we will specify the gradient boosted model with:

- Poisson distribution.  This is not the same as the usual 'family'.  It does
  not directly nominate a modelling family.  Instead it determines the **loss**
  function that is used to evaluate the tree splits.
- **var.monotone**: indicates which of the predictors should be constrained to
  monotonic trends (0: not monotonic, 1: positive monotonic, -1: negative
  monotonic)
- **n.trees**: the total number of trees to sequentially fit
- **interaction.depth**: the number of splits within each tree.  This roughly
  equates to the order of interactions.  The higher the number, the more complex
  the interactions can be, however, the faster the potential rate of learning
- **bag.fraction**: random fraction of data used to fit a single tree
- **shrinkage**: rate of learning.  The smaller the value, the slower the
  learning rate.  Typically, values should be between 0.001 and 0.01.
- **train.fraction**: the proportion of the data used to train the tree. This is
  for the purpose of determining the balance between over and underfitting and
  evaluate the accuracy of the fit, however for the former purpose, this is a
  cruder approach than either out-of-bag or cross validation. Therefore, it is
  arguably better to set this fraction to 1 (use 100 percent of the data for
  training) and use either out-of-bag or cross validation and (as we have done
  here), withhold a fraction of the full data and perform our own accuracy
  checking.
- **cv.folds**: the number of cross validation folds.

The values of `n.trees`, `interaction.depth` and `shrinkage` used below are
purely based on what are typically good starting points.  Nevertheless, they
will be data set dependent.  We will start off with those values and then evaluate
whether they are appropriate.

## Fit model

```{r fitModel1, results='hide', eval=TRUE, mhidden=TRUE, cache=TRUE}
newman_gbm <- gbm(fish_rich ~ fReef_complexity +
                   No_tall_corals +
                   No_corals +
                   Sponge_max_height +
                   Octocoral_max_height +
                   Slope_angle +
                  fCountry,
  data = newman,
  distribution = "poisson",
  n.trees = 10000,
  var.monotone = c(0, 1, 1, 1, 0, 1, 0),
  interaction.depth = 5,
  shrinkage = 0.001,
  bag.fraction = 0.5,
  train.fraction = 1,
  n.minobsinnode = 10,
  verbose = FALSE,
  cv.folds = 10,
  n.cores = 1
)
```

## Determine optimum number of trees

We will now determine the optimum number of trees estimated to be required in
order to achieve a balance between bias (biased towards the exact observations)
and precision (variability in estimates). Ideally, the optimum number of trees
should be close to 1000. If it is much less (as in this case), it could imply
that the tree learned too quickly. On the other hand, if the optimum number of
trees is very close to the total number of fitted trees, then it suggests that
the optimum may not actually have occured yet and that more trees should be used
(or a faster learning rate).

```{r fitModel2, results='markdown', eval=TRUE, mhidden=TRUE}
(best.iter <- gbm::gbm.perf(newman_gbm, method='OOB'))
(best.iter <- gbm::gbm.perf(newman_gbm, method = 'cv'))
#OR
#(best_iter <- gbm3::gbmt_performance(newman_gbm, method = "cv"))
#plot(best_iter)
```

:::


<!-- END_PRIVATE-->

# Explore relative influence

<!-- START_PRIVATE-->

If a predictor is an important driver of the patterns in the response, then many
of the tree splits should feature this predictor.  It thus follows that the
number of proportion of total splits that features each predictor will be a
measure of the relative influence of each of the predictors.

```{r relativeInfluence1, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=7, fig.height=10}
#| dependson: fitModel3
summary(newman_gbm, n.trees = best.iter)
## summary(newman_gbm, num_trees = best.iter)
100 / 7

```

**Conclusions:**

- the most influential predictor is Country.  This is a categorical
  predictor that is going to act as a proxy for all the various
  characteristics that differ between the different countries.  Corals
  do not respond directly to countries, rather they are likely to be
  responding to the underlying differences in characteristics.
- as a categorical predictor (and one that is going to be a proxy for
  a large number of characteristics), Country will outcompete the
  individual characteristics
- if we want to explore the relative impact of the characteristics,
  then we need to exclude Country.


```{r fitModel2a, results='hide', eval=TRUE, mhidden=TRUE, cache=TRUE}
newman_gbm <- gbm(fish_rich ~ fReef_complexity +
                   No_tall_corals +
                   No_corals +
                   Sponge_max_height +
                   Octocoral_max_height +
                   Slope_angle,
  data = newman,
  distribution = "poisson",
  n.trees = 10000,
  var.monotone = c(0, 1, 1, 1, 0, 1),
  interaction.depth = 5,
  shrinkage = 0.001,
  bag.fraction = 0.5,
  train.fraction = 1,
  n.minobsinnode = 10,
  verbose = FALSE,
  cv.folds = 10,
  n.cores = 1
)
```

```{r fitModel2b, results='markdown', eval=TRUE, mhidden=TRUE}
(best.iter <- gbm::gbm.perf(newman_gbm, method='OOB'))
(best.iter <- gbm::gbm.perf(newman_gbm, method = 'cv'))
#OR
#(best_iter <- gbm3::gbmt_performance(newman_gbm, method = "cv"))
#plot(best_iter)
```

```{r relativeInfluence1b, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=7, fig.height=10}
#| dependson: fitModel3
summary(newman_gbm, n.trees = best.iter)
100 / 6

```




<!-- END_PRIVATE-->

# Explore partial effects

<!-- START_PRIVATE-->

::: {.panel-tabset}

## Autoplot

```{r partialEffects1, results='markdown', eval=TRUE, mhidden=TRUE, fig.width=4, fig.height=4}
#| dependson: fitModel3

attr(newman_gbm$Terms,"term.labels")

interact.gbm(newman_gbm, newman, c(1,5), n.tree=best.iter)

attr(newman_gbm$Terms,"term.labels")
plot(newman_gbm, 5, n.tree = best.iter)
plot(newman_gbm, 1, n.tree = best.iter, log = 'x')
## plot(newman_gbm, 7, num.tree = best.iter)
## plot(newman_gbm, 1, num.tree = best.iter, log = 'x')
newman_gbm |>
    pdp::partial(pred.var='fReef_complexity',
                 n.trees = best.iter,
                 type = "regression",
                 recursive = FALSE,
                 ice = TRUE,
                 inv.link = exp) |>
  autoplot()
## newman_gbm |>
##     pdp::partial(pred.var='fCountry',
##                  n.trees = best.iter,
##                  type = "regression",
##                  recursive = FALSE,
##                  ice = TRUE,
##                  inv.link = exp) |>
##   autoplot()

newman_gbm |>
    pdp::partial(pred.var='Octocoral_max_height',
                 n.trees = best.iter,
                 type = "regression",
                 recursive = FALSE,
                 ice = TRUE,
                 inv.link = exp) |>
  autoplot()
```

:::
<!-- END_PRIVATE-->

# Bootstrapping

<!-- START_PRIVATE-->
```{r bootstrapping, results='markdown', eval=TRUE, mhidden=TRUE, fig.width = 7, fig.height = 5}

## define a function that resamples the data
resample_data <- function(dat) {
  dat <- dat |> sample_n(size = n(), replace = TRUE)
  return(dat)
}

fit_tree <- function(dat) {
  mod_gbm <- gbm(fish_rich ~ fReef_complexity +
                      No_tall_corals +
                      No_corals +
                      Sponge_max_height +
                      Octocoral_max_height +
                      Slope_angle,
                      ## fCountry,
                      data = dat,
                      distribution = "poisson",
                      n.trees = 10000,
                      var.monotone = c(0, 1, 1, 1, 0, 1),
                      interaction.depth = 5,
                      shrinkage = 0.001,
                      bag.fraction = 0.5,
                      train.fraction = 1,
                      n.minobsinnode = 10,
                      verbose = FALSE,
                      cv.folds = 10,
                      n.cores = 1
  )
  return(mod_gbm)
}

## newdata <- with(newman,
##   expand.grid(
##     Octocoral_max_height = seq(min(Octocoral_max_height), max(Octocoral_max_height), len = 100),
##     fCountry = levels(fCountry),
##     fReef_complexity =  NA,
##     No_tall_corals = NA,
##     No_corals = NA,
##     Sponge_max_height =  NA,
##     Slope_angle = NA)
## ) |>
##   mutate(fCountry = factor(fCountry))

newdata <- with(
  newman,
  expand.grid(
    fReef_complexity = levels(fReef_complexity),
    Octocoral_max_height = seq(min(Octocoral_max_height), max(Octocoral_max_height), len = 100),
    ## fCountry = levels(fCountry),
    No_tall_corals = NA,
    No_corals = NA,
    Sponge_max_height = NA,
    Slope_angle = NA
  )
) |>
  mutate(
    ## fCountry = factor(fCountry),
    fReef_complexity = factor(fReef_complexity))

nBoot <- 10
pred.list <- vector('list', nBoot)
rel.inf.list <- vector('list', nBoot)

for (i in 1:nBoot) {
  print(paste0('Boot number: ', i))
  ## Resample the data
  dat <- resample_data(newman)
  ## Fit the tree
  mod <- fit_tree(dat)
  ## Determine the best number of trees
  best.iter <- gbm.perf(mod, method = 'cv')
  ## predict based on shell weight
  fit <- predict(mod, newdata = newdata, n.trees = best.iter)
  pred.list[[i]] <- data.frame(newdata, Boot = i, Fit = fit)
  ## relative influence
  rel.inf.list[[i]] <- summary(mod, n.trees = best.iter)
}

newman.fit <- do.call('rbind', pred.list)
## newman.fit <- newman.fit |>
##     group_by(Octocoral_max_height, fCountry) |>
##     ggdist::median_hdci(Fit)
newman.fit <- newman.fit |>
    group_by(fReef_complexity, Octocoral_max_height) |>
    tidybayes::median_hdci(Fit)

## g1 <-
##   newman.fit |> ggplot(aes(y=Fit, x = Octocoral_max_height,
##   fill = fCountry, colour = fCountry)) +
##   geom_ribbon(aes(ymin=.lower, ymax=.upper), alpha=0.3, color=NA) +
##   geom_line() +
##   scale_fill_viridis_d() +
##   scale_colour_viridis_d() +
##   scale_x_log10() +
##   theme_classic()
g1 <-
  newman.fit |>
  ggplot(aes(y=Fit, x = Octocoral_max_height, color = fReef_complexity)) +
  ## fill = fCountry, colour = fCountry)) +
  geom_ribbon(aes(ymin =.lower, ymax =.upper, fill = fReef_complexity),
    colour = NA, alpha = 0.3) +
  geom_line() +
  scale_colour_viridis_d("Reef\ncomplexity") +
  scale_fill_viridis_d("Reef\ncomplexity") +
  theme_classic()

rel.inf<- do.call('rbind', rel.inf.list)
rel.inf <- rel.inf |>
    group_by(var) |>
    tidybayes::median_hdci(rel.inf)

g2 <-
  rel.inf |>
  arrange(rel.inf) |>
  mutate(var =  factor(var, levels = unique(var))) |>
  ggplot(aes(y=var, x=rel.inf)) +
    geom_vline(xintercept=12.5, linetype='dashed') +
    geom_pointrange(aes(xmin=.lower, xmax=.upper)) +
    theme_classic()

g2 + patchwork::inset_element(g1, left=0.4, bottom=0.01, right=1, top=0.4)
g1 + patchwork::inset_element(g2, left=0.01, bottom=0.01, right=0.5, top=0.35)
```

```{r}
library(randomForest)
newman.rf <- randomForest(
  fish_rich ~ fReef_complexity +
                      No_tall_corals +
                      No_corals +
                      Sponge_max_height +
                      Octocoral_max_height +
                      Slope_angle,
                      ## fCountry,
                       data=newman, importance=TRUE,
                       ntree=1000)
newman.imp <- randomForest::importance(newman.rf)


## Rank by either:
## *MSE (mean decrease in accuracy)
## For each tree, calculate OOB prediction error.
## This also done after permuting predictors.
## Then average diff of prediction errors for each tree
## *NodePurity (mean decrease in node impurity)
## Measure of the total decline of impurity due to each
## predictor averaged over trees
100*newman.imp/sum(newman.imp)
varImpPlot(newman.rf)
## use brute force
newman.rf |>
    pdp::partial("No_tall_corals") |>
    autoplot()
```
